\documentclass[]{statthesisMEMOIR}

\title{The Subordination of Women and National Health Outcomes \titlebreak A Bayesian Hierarchical Regression Model Approach}
\author{Kaylee Britton Hodgson}
\chair{Richard Warr}
\committee{Gilbert Fellingham}
\committeetwo{Garritt Page}
\typeofwork{project} % project or thesis
\graddate{April 2019}
\copyrightyear{2019}
\keywords{hierarchical model, bayesian, multivariate, regression, missing data}

\abstract{How does the treatment of women impact health outcomes? Specifically, how are the health of women, children, and the general public in a country impacted by how systematically and severely women are subordinated in the household in that country? I use a Bayesian hierarchical regression model to test whether poor measures of health outcomes are associated with higher levels of subordination of women. The Bayesian hierarchical regression model is proposed over using multivariate or simple regression analyses because of 1) the non-random missingness in the data, and 2) the relationship between the health outcomes. Before implementing the proposed Bayesian hierarchical regression model, I use a simulation study to test it against these other models. I find that overall the Bayesian hierarchical model tends to estimate with less bias and more accuracy than the other models. \titlebreak Using the hierarchical model, I find that higher levels of subordination of women in the household are significantly and consistently associated with worse health outcomes for women and children, and that while the results are less consistent for overall health, that the same relationship exists in most of those models. I propose that political and community leaders working to improve health outcomes place higher focus on the empowerment of women to achieve their objectives.}

\acknowledgments{I would first like to thank my project chair Dr. Richard Warr of the Statistics Department at Brigham Young University. Dr. Warr mentored and guided my work on this project, while also listening to and encouraging my ideas for it. He was always willing to take time to teach me new concepts, and to help me improve the research and writing. \titlebreak  I would also like to thank Dr. Valerie M. Hudson of the Department of International Affairs at Texas A\&M University, Dr. Perpetua Lynne Nielsen of the Department of Statistics at Brigham Young University, and Dr. Donna Lee Bowen of the Department of Political Science at Brigham Young University for their mentorship and their large contributions to this project. The dataset used in this analysis was originally conceptualized by them, and they allowed me to be an active participant in building the dataset. I especially appreciate the countless hours they spent with me teaching me about the theory behind the dataset and analysis, and for the analysis and writing opportunities they gave me that better prepared me for completing this project. \titlebreak  I would like to acknowledge and thank Dr. Gilbert Fellingham and Dr. Garritt Page of the Statistics Department at Brigham Young University, as committee members for this project. I am very grateful for both of their valuable guidance and comments on this project. \titlebreak  Finally, I would like to profoundly thank my husband, James, for the consistent support, love, and encouragement through this degree, and specifically in working on this project. His sacrifices for, and emphatic encouragement of, my ambitions have made this possible. Thank you.} 

\begin{document}

\input Chapter1.tex		% file with Chapter 1 contents

\input Chapter2.tex		% file with Chapter 2 contents

\input Chapter3.tex		% file with Chapter 3 contents

\nocite{*}		       
\bibliography{bibliography}	

\end{document}	
